---
title: "SparseMSE Generalization"
author: "Mila Madiot"
date: "2026-06-16"
output: html_document
---

```{r}
library(SparseMSE)
library(stringr)
library(combinat)
```


## Simulating the data
```{r}

get_prob= function(age, famous,sex, sample){
  base_prob = 0.1
  
  if(sex==1 & sample==3){base_prob = base_prob + 0.15}
  
  base_prob = base_prob + 0.15*(famous==0)
  
  #if(age==2){base_prob = base_prob + 0.1}
  
  #if((age==1)|(age==3)){base_prob = base_prob - 0.1}
  
  pmax(pmin(base_prob, 1), 0)
}

colonI <- function(formula_X) {
  var_pattern <- "[A-Za-z][0-9]*"
  
  # 4th way interactions
  if(str_count(formula_X, ":")==3){
    formula_str = gsub(
      paste0("(", var_pattern, "):(", var_pattern, "):(", var_pattern, "):(",var_pattern,")"),
      "I(\\1*\\2*\\3*\\4)",
      formula_X
    )
  }
  
  # 3rd way interactions
  if(str_count(formula_X, ":")==2){
    formula_str = gsub(
      paste0("(", var_pattern, "):(", var_pattern, "):(", var_pattern, ")"),
      "I(\\1*\\2*\\3)",
      formula_X
    )
  }
  
  # pairs
  if(str_count(formula_X, ":")==1){
    formula_str = gsub(
      paste0("(", var_pattern, "):(", var_pattern, ")"),
      "I(\\1*\\2)",
      formula_X
    )
  }
  
  if(str_count(formula_X, ":")==0){
    formula_str = formula_X
  }
  
  return(formula_str)
}

insert_colon = function(x){
  var = regmatches(x, gregexpr("[A-Za-z][0-9]*", x))[[1]]
  paste0(var, collapse = ":")
}
```

```{r}
set.seed(42)
popSize = 20000
Nsamples = 3

famous   = sample(c(0,1), popSize, replace=TRUE, prob=c(0.4,0.6)) # 1 for famous
  sex    = sample(c(0,1), popSize, replace=TRUE, prob=c(0.5,0.5)) # 1 for woman, 2 for man
  age    = sample(c(0,1,2), popSize, replace=TRUE, prob=c(0.2,0.6,0.2))
  # created the population
  
  Samples = matrix(NA, popSize, Nsamples)
  inclusionProbs = matrix(NA, popSize, Nsamples)
  
  for (s in 1:Nsamples){
    inclusionProbs[,s] = mapply(get_prob,age=age, famous = famous,sex= sex, MoreArgs=list(sample=s))
    Samples[,s] = rbinom(popSize, 1, inclusionProbs[,s])
  }
  colnames(Samples) = c("G","W","Z")
  freq = 1
  dfile = cbind(Samples,age, famous, sex)
  colnames(dfile) = c("G","W","Z","B", "F","S")
  dfile = as.data.frame(dfile)
  tb = as.data.frame(table(dfile))
  B2 = model.matrix(~B, tb)[,2]
  B3 = model.matrix(~B, tb)[,3]
  names(B2) = NULL
  names(B3) = NULL
  tb$B = NULL
  tb$B2 = B2
  tb$B3 = B3
  tb = tb[, c("G","W","Z","F","S","B2","B3","Freq")]
  
  for (col in c('G','W','Z','F','S','B2','B3')){
    tb[[col]] = as.numeric(tb[[col]])
  }
  for (col in c('G','W','Z','F','S')){
    for (row in 1:nrow(tb)){
      if(tb[[col]][row] == 1) tb[[col]][row] = 0 else tb[[col]][row] = 1
    }
  }
  
  null_obs = sum(tb[(tb[,1] == 0) & (tb[,2] == 0) & (tb[,3] == 0),]$Freq)
  tb[tb[,1] == 0 & tb[,2] == 0 & tb[,3] == 0,]$Freq = 0
  
  tb
```


## Full 3rd order model

```{r}
me = c("G","W","Z","F","S","B2","B3")
twoe   = unlist(lapply(combn(me,2,simplify=FALSE),paste0, collapse = ""))
twoe = twoe[-which(twoe == "B2B3")]
threee = unlist(lapply(combn(me,3,simplify=FALSE),paste0, collapse = ""))
threee = threee[-which(threee %in% c("GWZ","GB2B3","WB2B3","ZB2B3","FB2B3","SB2B3"))]
```

```{r}
three_formula = sapply(c(me,twoe,threee), insert_colon)
three_model   = as.formula(paste("Freq ~",paste(three_formula, collapse=" + ")))
  
formula_X = strsplit(as.character(three_model), " ")[[3]]
formula_X = formula_X[!grepl("+", formula_X, fixed = TRUE)] 
formula_X = formula_X[!grepl("*", formula_X, fixed = TRUE)] 
  
formula_X = sapply(formula_X, colonI)
formula_X = paste0("Freq~", paste0(c(formula_X), collapse="+"), collapse="")
  
l <- rep(list(0:1), 7)
X = as.data.frame(cbind(rep(1,nrow(as.matrix(expand.grid(l))[-1,])),
                        as.matrix(expand.grid(l))[-1,]))
colnames(X) = c("Freq",me)
X = as.matrix(model.frame(as.formula(formula_X), data = X))
dim(X)
```

# Identifiability

## Function

```{r}
model_to_X = function(model, nb_var = 7, tb){
  # model given without "Freq ~ "
  
  formula_X = strsplit(as.character(model), " ")[[3]]
  formula_X = formula_X[!grepl("+", formula_X, fixed = TRUE)] 
  formula_X = formula_X[!grepl("*", formula_X, fixed = TRUE)] 
  
  formula_X = sapply(formula_X, colonI)
  formula_X = paste0("Freq~", paste0(c(formula_X), collapse="+"), collapse="")
  
  l <- rep(list(0:1), nb_var)
  X = as.data.frame(cbind(rep(1,nrow(as.matrix(expand.grid(l))[-1,])),
                        as.matrix(expand.grid(l))[-1,]))
  colnames(X) = c("Freq",me)
  
  # empty overlaps that could have been observed
  # but here it is not the combination of variables that give 0 regardless of other variable
  # for example : eo can indicate that GZWF has 0 counts
  # but still GZWFS can have some counts
  out2 = tb[(tb$Freq == 0)&((tb$G!=0)|(tb$W!=0)|(tb$Z!=0)),]
  out3 = matrix(NA, nrow=nrow(out2), ncol= nb_var)
  for (row in 1:nrow(out2)){
    param = (out2[row,]==1)
    for (col in 1:nb_var){
      if(param[col]) out3[row,col] = colnames(param)[col] else out3[row,col] = ""
    }
  }
  eo = apply(out3, 1 , paste0, collapse = "")
  
  # new formulation of the model that takes out the coeff without overlaps if necessary
  pattern = character(length(eo))
  for (pat in 1:length(eo)){
  tokens = unlist(regmatches(eo[pat], gregexpr("[A-Z][0-9]*", eo[pat])))
  joined = paste(tokens, collapse = "\\*")
  pattern[pat] = paste0("\\+I\\(", joined, "\\)")
  formula_X = gsub(pattern[pat], "", formula_X) # we take it out
  }
  
  X = as.matrix(model.frame(as.formula(formula_X), data = X)) # based on the new formula_X
  
  # take out non-existing terms :
  out = which((X[,2]==0)&(X[,3]==0)&(X[,4]==0))
  X_upd = X[-out,]
  X_upd = as.data.frame(X_upd)
  X_upd = X_upd[-which((X_upd$B2==1)&(X_upd$B3==1)),]
  X_mat = as.matrix(X_upd)
  
  return(list(X_mat = X_mat, emptyoverlaps = eo, formula_X = formula_X))
}
```
Here the function is adapted to the names of the dummies.

## Test

```{r}
X_info = model_to_X(three_model, nb_var = 7, tb=tb)
dim(X_info$X_mat)[2] - qr(X_info$X_mat)$rank
```

If the result is 0, the model is identifiable.

# Existence

```{r}
counts = numeric()
X_upd = as.data.frame(X_info$X_mat)
for (row in 1:nrow(X_upd)){ # and not nrow(tb) because it includes more rows
  tbrow = 1
  found = 0
  while(found == 0){ # we test until we find the match
    if(mean(tb[tbrow,-8] == X_upd[row,2:8])==1){
      counts[row] = tb[tbrow,8]
      found = 1
    } else {
      tbrow = tbrow + 1
    }
  }
}
length(counts)
```
The column indices should be adapted so that it compares the main effects columns (without the "Freq" column).
```{r}
check_exist_adapt = function(X_mat, counts){
  amat = t(X_mat)
  tvec = t(X_mat) %*% counts
  tt = tvec
  npar = dim(amat)[1]
  nobs = dim(amat)[2]
  f.obj = c(rep(0, nobs), 1)
  const.rhs = c(tt, rep(0, nobs))
  amat1 = cbind(amat, rep(0, npar))
  amat2 = cbind(diag(nobs), rep(-1, nobs))
  const.mat = rbind(amat1, amat2)
  const.dir = c(rep("=", npar), rep(">=", nobs))
  zlp = lpSolve::lp("max", f.obj, const.mat, const.dir, const.rhs)
  return(list(s = zlp$objval))
}
check_exist_adapt(X_info$X_mat, counts)
```

If s=0, the estimate is not finite. If s is strictly positive, the estimate exists and is finite.
